### 2022-06-28 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@ec3555b2e83eb8fec7839bfb58e33f91be22e17e 🚀

The Geographic Mosaic Theory of Coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic interactions.\textsuperscript{7} The GMTC
posits that coevolutionary processes among pairs\textsuperscript{8} or
complexes\textsuperscript{9} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanisms,
giving rise to fragmented systems with different ecologies over large
spatial extents.\textsuperscript{10} The GMTC predicts a spatial
fragmentation of coevolutionary dynamics under the joint action of three
processes:\textsuperscript{11} coevolutionary hot- and coldspots, which
appear when the intensity of \emph{interaction} (in terms of reciprocal
fitness consequences) varies spatially; selection mosaics, wherein the
intensity of \emph{selection} varies across space, driven by both the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex) and the local favorability of the
environment;\textsuperscript{12} and trait remixing, which occurs when
coevolutionary dynamics change when community-level \emph{functional
traits} change through meta-community dynamics.

@@

\hypertarget{coevolution-informed-emergence-risk-is-different-in-space}{%
\subsection{Coevolution-informed emergence risk is different in
space}\label{coevolution-informed-emergence-risk-is-different-in-space}}

Both the diverse origins of betacoronavirus spillover events and the
existence of well defined co-phylogenetic regions suggest that the
system is spatially fragmented enough to create risks of emergence that
are different in their amplitude and in the coevolutionary processes
underlying them. Therefore, we turn to the Geographic Mosaic Theory of
Coevolution to provide an assesment of risk accounting for these
processes. In fig.~\ref{fig:trivariate}, we overlapped three components
of spillover risk: viral sharing (high viral sharing suggests slower,
diffuse coevolution); host phylogenetic diversity (high diversity
represents different evolutionary histories, suggesting more variation
in immune strategies); host community uniqueness (high uniqueness
suggests more potential for viruses to be exposed to novel host traits).
This approach leads to the definition of broad biogeographic regions of
risk, where close colors represent similar risks, and paler pixels
represent overall higher risk (see Methods). These regions do not neatly
overlap with those defined in fig.~\ref{fig:richness} or
fig.~\ref{fig:biogeo}, reinforcing the notion that local-scale
coevolutionary mosaics exist within co-phylogenetic regions.

@@

Under this framework, other regions (where high viral sharing dominates
the dynamics; see \emph{e.g.} Latin America, Eurasia above a northing of
30) represent a lower risk of emergence. Nevertheless, areas of high
host uniqueness coupled with high viral sharing(red-to-pink in
fig.~\ref{fig:trivariate}) could provide future hotspots of
\emph{Betacoronavirus} emregence risk through the suddent divergence of
currently diffuse coevolutionary dynamics; this is a likely scenario
knowing that can be facilitated by codivergence followed by
recombination. Madagascar, where most bat species are endemic following
evolutionary divergence from sister species in both African and Asian
continents,\textsuperscript{23} is one such potential future hotspot.
Indeed, recent surveillance\textsuperscript{24} has identified a novel
\emph{Betacoronavirus} (in the subgenus \emph{Nobecovirus}) in
Madagascar-endemic pteropid bat species (\emph{Pteropus rufus},
\emph{Rousettus madagascariensis}), emphasizing strong proof of
principle in model predictions.

@@

The relationship between the underlying pathogen pool and emergence risk
is mediated by both human-wildlife interfaces (the probability of
spillover) and opportunities for onward transmission (the probability
that spillovers become epidemics)\textsuperscript{1}. As a proxy for
both, we finally overlaid the risk component from the composite map (see
above) with the proportion of built land, as a proxy for a mix of
habitat disturbance, potential for bat synanthropy or contact with
bridge hosts like livestock,\textsuperscript{25,26} and human population
density and connectivity\textsuperscript{1,27,28}
(fig.~\ref{fig:compound}). Accounting for these factors, most of South
America and Europe are at comparatively lower risk, as--although densely
populated--settlements tend to be in areas with lower potential risk.
Conversely, regions like Malaysia and the northern coast of Australia
have a high evolutionary risk component, but should represent a
relatively lower effective risk due to low human density. However,
southeast Asia, the Indian subcontinent, and scattered hotspots in
sub-Saharan Africa are at high risk due to the overlap between human
populations and natural opportunities for cross-species transmission of
betacoronaviruses.

@@

Reassuringly, these predictions correspond to the geographic origins of
the three bat-origin coronaviruses that have recently emerged in human
populations. While available information puts the spillover of
SARS-CoV-2 in a live animal market in Wuhan, China, the ultimate origin
of the virus is almost certainly in a divergent lineage of
sarbecoviruses from the Indochinese peninsula that was poorly
characterized prior to the pandemic.\textsuperscript{29--31} Similarly,
the SARS-CoV outbreak began in Guangdong province in 2002, reaching
humans through small carnivore bridge hosts, but was eventually traced
back to a set of likely progenitor viruses found in cave-dwelling
horseshoe bats in Yunnan province;\textsuperscript{32} nearby, antibody
evidence has indicated human exposure to SARS-like
viruses.\textsuperscript{33} MERS-CoV was originally detected in Saudi
Arabia, accompanied by a nearly identical virus sequenced from an
Egyptian tomb bat (\emph{Taphozous perforatus}),\textsuperscript{34} but
is widespread in camels in East Africa and the Middle East, and may have
reached its bridge host decades earlier than originally
supposed;\textsuperscript{35} as a result, the geography of the original
bat-to-camel transmission is still widely regarded as uncertain. All of
these are broadly consistent with the risk factors we identify. Notably,
India and west Africa are additional hotspots that have yet to
experience the emergence of a bat coronavirus into human populations,
but may still be at risk---particularly given known gaps in bat
surveillance,\textsuperscript{20} and a dense population in both regions
with global connectivity. In any of these regions, surveillance on viral
reservoirs can be paired with targeted monitoring of high-risk human
populations (i.e., those with regular wildlife
contact)\textsuperscript{36} for maximum impact.

@@

Bats are important reservoir hosts for different classes of
microorganisms, many of which a threat to human
health.\textsuperscript{37,38} Chiropterans emerged around 64 million
years ago and are one of the most diverse mammalian orders, with an
estimated richness of more than 1400 species.\textsuperscript{39,40}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to human.\textsuperscript{41} For example,
bats are an essential component of many seed-dispersal
networks.\textsuperscript{42} Over two-thirds of bats are know to be
either obligate or facultative insectivores, therefore actively
contributing for agricultural pest control,\textsuperscript{43,44} and
vectors of pathogens that put a risk on human
health.\textsuperscript{45,46} Because bats are globally distributed and
have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the contemporary distribution
of coevolutionary processes between bats and the pathogens they host.
Not all areas in which bats, viruses, and human are co-occuring are
facing a risk of spillover towards human populations, and the areas in
which this risk exist may not be facing risks of the same nature and
magnitude.

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{18} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{19} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{13} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{47} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{19,20} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{48}

The diversity and diversification potential of bats responds to
anthropogenic factors others than shifting climates.\textsuperscript{49}
Land use changes could significantly decrease bat suitability, notably
through effects on diet and availability of
habitats.\textsuperscript{50} As our results establish that the
diversification of bats betacoronaviruses happens on top of processes
affecting hosts, biogeographic variation in human population density and
anthropogenic disturbances may feed into co-evolutionary dynamics.
Increase in humans-hosts contacts also increase the risk of emergence of
novel diseases,\textsuperscript{51} so does the changes in landscape
connectivity at local/regional scales.\textsuperscript{52} This
represents a challenge for both conservation strategies and disease
ecology: some areas can a high emergence risk and more potential for the
acquisition of zoonotic viruses through bat-human
encounters.\textsuperscript{53} In particular, the challenge ahead lies
in the need to quantify actual exposure (and risk) accounting for
several transmission scenarios, including both direct and indirect bat -
human interactions, and feeding back into the provision of ecosystem
services by bats.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{61} This network stores pairwise values of
viral community similarity. To project viral sharing values into a
single value for every pixel, we averaged the pairwise scores. High
values of the average sharing propensity means that this specific extant
bat assemblage is likely to be proficient at exchanging viruses.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{62} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{63} in that this allows multiple
independent host-virus coevolutionary dynamics to take place in the same
location. In the colorimetric space, this correspond to yellow --
because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{64} we measured the risk level by
calculating the angular distance of the hue of each pixel to a reference
value of 60 (yellow), and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent'',
``mutant'', ``GFP'', and ``recombinant''. We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using
MAFFT\textsuperscript{65} v1.4.0 (Algorithm FFT-NS-2, Scoring matrix
200PAM / k=2, gap open penalty 1.53m offset value 0.123) and a maximum
likelihood tree reconstructed in IQ-TREE\textsuperscript{66} v1.6.12
with ModelFinder\textsuperscript{67} ultrafast bootstrap
approximation\textsuperscript{68} with a general time reversible model
with empirical base frequencies and the 5-discrete-rate-category
FreeRaye model of nucleotide substitution (GTR+F+R5).

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness\textsuperscript{69} for each of the viruses in the tree,
then averaged these at the bat species level, projected these values
onto their geographic distributions, and averaged across every bat found
in a given pixel. As such, this can be thought of as a map of the mean
evolutionary distinctiveness of the known viral community believed to be
associated with a particular subset of bats present.

@@

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from,\textsuperscript{70,71} who
proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map. Here, we build on this idea with an
entirely novel methodology. First, we measure the phylogenetic distance
between the different viruses in the betacoronaviruses tree by using the
cophenetic function in \texttt{ape};\textsuperscript{72} subsequently,
we take a principal components analysis of that distance matrix (readily
interchangeable for NMDS in this case) to project the viral tree into an
n-dimensional space. We then take the first two principal components
and, as with the evolutionary distinctiveness analysis, aggregated these
to a mean host value and projected them using a four-color bivariate
map.

@@

\leavevmode\hypertarget{ref-Shi2014DeeDiv}{}%
\CSLLeftMargin{23. }
\CSLRightInline{Shi, J. J. \emph{et al.} A Deep Divergence Time between
Sister Species of Eidolon (Pteropodidae) with Evidence for Widespread
Panmixia. \emph{Acta Chiropterologica} \textbf{16}, 279--292 (2014).}

\leavevmode\hypertarget{ref-Kettenburg2022FulGen}{}%
\CSLLeftMargin{24. }
\CSLRightInline{Kettenburg, G. \emph{et al.} Full Genome Nobecovirus
Sequences From Malagasy Fruit Bats Define a Unique Evolutionary History
for This Coronavirus Clade. \emph{Frontiers in Public Health}
\textbf{10}, (2022).}

\leavevmode\hypertarget{ref-Rulli2021LanCha}{}%
\CSLLeftMargin{25. }
\CSLRightInline{Rulli, M. C., D'Odorico, P., Galli, N. \& Hayman, D. T.
Land-use change and the livestock revolution increase the risk of
zoonotic coronavirus transmission from rhinolophid bats. \emph{Nature
Food} \textbf{2}, 409--416 (2021).}

\leavevmode\hypertarget{ref-Cui2019OriEvo}{}%
\CSLLeftMargin{26. }
\CSLRightInline{Cui, J., Li, F. \& Shi, Z.-L. Origin and evolution of
pathogenic coronaviruses. \emph{Nature Reviews Microbiology}
\textbf{17}, 181--192 (2019).}

\leavevmode\hypertarget{ref-Muylaert2022PreFut}{}%
\CSLLeftMargin{27. }
\CSLRightInline{Muylaert, R. L. \emph{et al.} Present and future
distribution of bat hosts of sarbecoviruses: Implications for
conservation and public health. \emph{Proceedings of the Royal Society
B: Biological Sciences} \textbf{289}, 20220397 (2022).}

\leavevmode\hypertarget{ref-Hassell2017UrbDis}{}%
\CSLLeftMargin{28. }
\CSLRightInline{Hassell, J. M., Begon, M., Ward, M. J. \& Fèvre, E. M.
Urbanization and disease emergence: Dynamics at the
wildlife--livestock--human interface. \emph{Trends in ecology \&
evolution} \textbf{32}, 55--67 (2017).}

\leavevmode\hypertarget{ref-Worobey2022HuaMar}{}%
\CSLLeftMargin{29. }
\CSLRightInline{Worobey, M. \emph{et al.} The Huanan market was the
epicenter of SARS-CoV-2 emergence. \emph{Zenodo} (2022)
doi:\href{https://doi.org/10.5281/zenodo.6299600}{10.5281/zenodo.6299600}.}

\leavevmode\hypertarget{ref-Temmam2022BatCor}{}%
\CSLLeftMargin{30. }
\CSLRightInline{Temmam, S. \emph{et al.} Bat coronaviruses related to
SARS-CoV-2 and infectious for human cells. \emph{Nature} \textbf{604},
330--336 (2022).}

>>>>

The geographic mosaic theory of coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic interactions.\textsuperscript{7} The GMTC
posits that coevolutionary processes among pairs\textsuperscript{8} or
complexes\textsuperscript{9} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanisms,
giving rise to fragmented systems with different ecologies over large
spatial extents.\textsuperscript{10} The GMTC predicts a spatial
fragmentation of coevolutionary dynamics under the joint action of three
processes:\textsuperscript{11} coevolutionary hot- and coldspots, which
appear when the intensity of \emph{interaction} (in terms of reciprocal
fitness consequences) varies spatially; selection mosaics, wherein the
intensity of \emph{selection} varies across space, driven by both the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex) and the local favorability of the
environment;\textsuperscript{12} and trait remixing, which occurs when
coevolutionary dynamics change when community-level \emph{functional
traits} change through meta-community dynamics.

@@

\hypertarget{the-coevolutionary-mosaic-generates-emergence-regimes}{%
\subsection{The coevolutionary mosaic generates emergence
regimes}\label{the-coevolutionary-mosaic-generates-emergence-regimes}}

The existence of well-defined cophylogenetic regions suggests that the
bat-betacoronavirus system is spatially fragmented enough to create
differential coevolutionary processes; in turn, the coevolutionary
mosaic contributes to heterogeneity in emergence risk. These ideas are,
respectively, supported by the existence of hotspots of viral uniqueness
and the diverse origins of human betacoronaviruses. Together, these
ideas point to a predictable relationship between host community
structure and coevolutionary pressure: phylogeographic structure in bat
hosts---and their diverse immune
strategies---\textsuperscript{23}creates a landscape of selective
pressure; the trajectory of viruses' coevolutionary response is, in
turn, constrained by the opportunities they have for either
coevolutionary specialization or diversification through host jumps and
recombination.

Based on the geographic mosaic theory of coevolution, we developed a
trivariate map of three facets of coevolutionary pressure (see Methods):
(1) \emph{host phylogenetic diversity}: a high diversity of evolutionary
histories should expose viruses to more variation in host immune traits;
(2) \emph{host community uniqueness}: coevolving with more unique host
communities should create more unique branches of viral evolution; and
(3) propensity for \emph{viral sharing}: frequent cross-species
transmission may act as a buffer on selective pressure, while lower
rates of exchange may enable more simultaneous trajectories of viral
specialization to coexist within a given community. We combine global
maps of all three to generate a map of coevolutionary regimes, where
close colors represent similar risks, and paler pixels represent overall
higher risk (fig.~\ref{fig:trivariate}). We find that these regions do
not neatly overlap with those defined in fig.~\ref{fig:richness} or
fig.~\ref{fig:biogeo}, reinforcing the notion that local-scale
coevolutionary mosaics can form within cophylogenetic regions.

@@

Under this framework, other regions (where high viral sharing dominates
the dynamics; see \emph{e.g.} Latin America, Eurasia above a northing of
30) represent a lower risk of emergence. Nevertheless, areas of high
host uniqueness coupled with high viral sharing(red-to-pink in
fig.~\ref{fig:trivariate}) could provide future hotspots of
\emph{Betacoronavirus} emregence risk through the suddent divergence of
currently diffuse coevolutionary dynamics; this is a likely scenario
knowing that can be facilitated by codivergence followed by
recombination. Madagascar, where most bat species are endemic following
evolutionary divergence from sister species in both African and Asian
continents,\textsuperscript{24} is one such potential future hotspot.
Indeed, recent surveillance\textsuperscript{25} has identified a novel
\emph{Betacoronavirus} (in the subgenus \emph{Nobecovirus}) in
Madagascar-endemic pteropid bat species (\emph{Pteropus rufus},
\emph{Rousettus madagascariensis}), emphasizing strong proof of
principle in model predictions.

@@

The relationship between the underlying pathogen pool and emergence risk
is mediated by both human-wildlife interfaces (the probability of
spillover) and opportunities for onward transmission (the probability
that spillovers become epidemics)\textsuperscript{1}. As a proxy for
both, we finally overlaid the risk component from the composite map (see
above) with the proportion of built land, as a proxy for a mix of
habitat disturbance, potential for bat synanthropy or contact with
bridge hosts like livestock,\textsuperscript{26,27} and human population
density and connectivity\textsuperscript{1,28,29}
(fig.~\ref{fig:compound}). Accounting for these factors, most of South
America and Europe are at comparatively lower risk, as--although densely
populated--settlements tend to be in areas with lower potential risk.
Conversely, regions like Malaysia and the northern coast of Australia
have a high evolutionary risk component, but should represent a
relatively lower effective risk due to low human density. However,
southeast Asia, the Indian subcontinent, and scattered hotspots in
sub-Saharan Africa are at high risk due to the overlap between human
populations and natural opportunities for cross-species transmission of
betacoronaviruses.

@@

Reassuringly, these predictions correspond to the geographic origins of
the three bat-origin coronaviruses that have recently emerged in human
populations. While available information puts the spillover of
SARS-CoV-2 in a live animal market in Wuhan, China, the ultimate origin
of the virus is almost certainly in a divergent lineage of
sarbecoviruses from the Indochinese peninsula that was poorly
characterized prior to the pandemic.\textsuperscript{30--32} Similarly,
the SARS-CoV outbreak began in Guangdong province in 2002, reaching
humans through small carnivore bridge hosts, but was eventually traced
back to a set of likely progenitor viruses found in cave-dwelling
horseshoe bats in Yunnan province;\textsuperscript{33} nearby, antibody
evidence has indicated human exposure to SARS-like
viruses.\textsuperscript{34} MERS-CoV was originally detected in Saudi
Arabia, accompanied by a nearly identical virus sequenced from an
Egyptian tomb bat (\emph{Taphozous perforatus}),\textsuperscript{35} but
is widespread in camels in East Africa and the Middle East, and may have
reached its bridge host decades earlier than originally
supposed;\textsuperscript{36} as a result, the geography of the original
bat-to-camel transmission is still widely regarded as uncertain. All of
these are broadly consistent with the risk factors we identify. Notably,
India and west Africa are additional hotspots that have yet to
experience the emergence of a bat coronavirus into human populations,
but may still be at risk---particularly given known gaps in bat
surveillance,\textsuperscript{20} and a dense population in both regions
with global connectivity. In any of these regions, surveillance on viral
reservoirs can be paired with targeted monitoring of high-risk human
populations (i.e., those with regular wildlife
contact)\textsuperscript{37} for maximum impact.

@@

Bats are important reservoir hosts for different classes of
microorganisms, many of which a threat to human
health.\textsuperscript{38,39} Chiropterans emerged around 64 million
years ago and are one of the most diverse mammalian orders, with an
estimated richness of more than 1400 species.\textsuperscript{40,41}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to human.\textsuperscript{42} For example,
bats are an essential component of many seed-dispersal
networks.\textsuperscript{43} Over two-thirds of bats are know to be
either obligate or facultative insectivores, therefore actively
contributing for agricultural pest control,\textsuperscript{44,45} and
vectors of pathogens that put a risk on human
health.\textsuperscript{46,47} Because bats are globally distributed and
have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the contemporary distribution
of coevolutionary processes between bats and the pathogens they host.
Not all areas in which bats, viruses, and human are co-occuring are
facing a risk of spillover towards human populations, and the areas in
which this risk exist may not be facing risks of the same nature and
magnitude.

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{18} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{19} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{13} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{48} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{19,20} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{49}

The diversity and diversification potential of bats responds to
anthropogenic factors others than shifting climates.\textsuperscript{50}
Land use changes could significantly decrease bat suitability, notably
through effects on diet and availability of
habitats.\textsuperscript{51} As our results establish that the
diversification of bats betacoronaviruses happens on top of processes
affecting hosts, biogeographic variation in human population density and
anthropogenic disturbances may feed into co-evolutionary dynamics.
Increase in humans-hosts contacts also increase the risk of emergence of
novel diseases,\textsuperscript{52} so does the changes in landscape
connectivity at local/regional scales.\textsuperscript{53} This
represents a challenge for both conservation strategies and disease
ecology: some areas can a high emergence risk and more potential for the
acquisition of zoonotic viruses through bat-human
encounters.\textsuperscript{54} In particular, the challenge ahead lies
in the need to quantify actual exposure (and risk) accounting for
several transmission scenarios, including both direct and indirect bat -
human interactions, and feeding back into the provision of ecosystem
services by bats.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{62} This network stores pairwise values of
viral community similarity. To project viral sharing values into a
single value for every pixel, we averaged the pairwise scores. High
values of the average sharing propensity means that this specific extant
bat assemblage is likely to be proficient at exchanging viruses.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{63} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{64} in that this allows multiple
independent host-virus coevolutionary dynamics to take place in the same
location. In the colorimetric space, this correspond to yellow --
because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{65} we measured the risk level by
calculating the angular distance of the hue of each pixel to a reference
value of 60 (yellow), and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent'',
``mutant'', ``GFP'', and ``recombinant''. We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using
MAFFT\textsuperscript{66} v1.4.0 (Algorithm FFT-NS-2, Scoring matrix
200PAM / k=2, gap open penalty 1.53m offset value 0.123) and a maximum
likelihood tree reconstructed in IQ-TREE\textsuperscript{67} v1.6.12
with ModelFinder\textsuperscript{68} ultrafast bootstrap
approximation\textsuperscript{69} with a general time reversible model
with empirical base frequencies and the 5-discrete-rate-category
FreeRaye model of nucleotide substitution (GTR+F+R5).

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness\textsuperscript{70} for each of the viruses in the tree,
then averaged these at the bat species level, projected these values
onto their geographic distributions, and averaged across every bat found
in a given pixel. As such, this can be thought of as a map of the mean
evolutionary distinctiveness of the known viral community believed to be
associated with a particular subset of bats present.

@@

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from,\textsuperscript{71,72} who
proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map. Here, we build on this idea with an
entirely novel methodology. First, we measure the phylogenetic distance
between the different viruses in the betacoronaviruses tree by using the
cophenetic function in \texttt{ape};\textsuperscript{73} subsequently,
we take a principal components analysis of that distance matrix (readily
interchangeable for NMDS in this case) to project the viral tree into an
n-dimensional space. We then take the first two principal components
and, as with the evolutionary distinctiveness analysis, aggregated these
to a mean host value and projected them using a four-color bivariate
map.

@@

\leavevmode\hypertarget{ref-Banerjee2020NovIns}{}%
\CSLLeftMargin{23. }
\CSLRightInline{Banerjee, A. \emph{et al.} Novel Insights Into Immune
Systems of Bats. \emph{Frontiers in Immunology} \textbf{11}, 26 (2020).}

\leavevmode\hypertarget{ref-Shi2014DeeDiv}{}%
\CSLLeftMargin{24. }
\CSLRightInline{Shi, J. J. \emph{et al.} A Deep Divergence Time between
Sister Species of Eidolon (Pteropodidae) with Evidence for Widespread
Panmixia. \emph{Acta Chiropterologica} \textbf{16}, 279--292 (2014).}

\leavevmode\hypertarget{ref-Kettenburg2022FulGen}{}%
\CSLLeftMargin{25. }
\CSLRightInline{Kettenburg, G. \emph{et al.} Full Genome Nobecovirus
Sequences From Malagasy Fruit Bats Define a Unique Evolutionary History
for This Coronavirus Clade. \emph{Frontiers in Public Health}
\textbf{10}, (2022).}

\leavevmode\hypertarget{ref-Rulli2021LanCha}{}%
\CSLLeftMargin{26. }
\CSLRightInline{Rulli, M. C., D'Odorico, P., Galli, N. \& Hayman, D. T.
Land-use change and the livestock revolution increase the risk of
zoonotic coronavirus transmission from rhinolophid bats. \emph{Nature
Food} \textbf{2}, 409--416 (2021).}

\leavevmode\hypertarget{ref-Cui2019OriEvo}{}%
\CSLLeftMargin{27. }
\CSLRightInline{Cui, J., Li, F. \& Shi, Z.-L. Origin and evolution of
pathogenic coronaviruses. \emph{Nature Reviews Microbiology}
\textbf{17}, 181--192 (2019).}

\leavevmode\hypertarget{ref-Muylaert2022PreFut}{}%
\CSLLeftMargin{28. }
\CSLRightInline{Muylaert, R. L. \emph{et al.} Present and future
distribution of bat hosts of sarbecoviruses: Implications for
conservation and public health. \emph{Proceedings of the Royal Society
B: Biological Sciences} \textbf{289}, 20220397 (2022).}

\leavevmode\hypertarget{ref-Hassell2017UrbDis}{}%
\CSLLeftMargin{29. }
\CSLRightInline{Hassell, J. M., Begon, M., Ward, M. J. \& Fèvre, E. M.
Urbanization and disease emergence: Dynamics at the
wildlife--livestock--human interface. \emph{Trends in ecology \&
evolution} \textbf{32}, 55--67 (2017).}

\leavevmode\hypertarget{ref-Worobey2022HuaMar}{}%
\CSLLeftMargin{30. }
\CSLRightInline{Worobey, M. \emph{et al.} The Huanan market was the
epicenter of SARS-CoV-2 emergence. \emph{Zenodo} (2022)
doi:\href{https://doi.org/10.5281/zenodo.6299600}{10.5281/zenodo.6299600}.}

\leavevmode\hypertarget{ref-Temmam2022BatCor}{}%
\CSLLeftMargin{31. }
\CSLRightInline{Temmam, S. \emph{et al.} Bat coronaviruses related to
SARS-CoV-2 and infectious for human cells. \emph{Nature} \textbf{604},
330--336 (2022).}
